Network Origins of Social Distinction

### 2025-09-20 manuscript.tex
9/13 changes
Updates from Overleaf

\citet[1263-4]{padgett1993robust} were the first to bring attention to how the cohesion of those around a broker can structure their own decisions. Cosimo de Medici, acting as a broker between two antagonistic groups, was constrained in the actions he could take because strong actions favoring one side would upset the other side. However, such a constraint was not an issue because Medici’s interests were in perpetuating his central role as broker, and, as the only source of interest, he was able to maintain the position. With regard to conceptualizing distinction, the theory of robust action puts forward that those at the center of political networks, as long as their interests derive solely from maintaining that position, engage in moderate strategies that do not overly antagonize either side. The decisions derive from ensuring the continuation of their position at the center of the network.

@@

The second term $\kappa_i$ in equation~\ref{eq:distinction} is a \textit{localized} induced centrality score in the sense of \citet{borgatti2020oxford-fa6}. According to \citet{everett2010induced-817}, an induced centrality is obtained whenever we calculate the difference that removing a node in the graph makes to some global property of the graph (a graph invariant). \citet{everett2010induced-817} suggest taking the sum of the centralities scores of the nodes in the node-deleted subgraph produced by removing a given node as that node's contribution to the centralities of other nodes in the graph, as a measure of that node's centrality. They propose examples extending this idea to the classical centrality measures like degree, closeness, and betweenness \citep{freeman79}.

@@

Triadic closure is an observed regularity in which a connection is likely to form between two actors who, although not originally connected, are first connected to a common alter (Simmel 1908). In essence, friends of friends have an increased likelihood of also becoming friends. The causes of this regularity, and the reasons for its occasional failure, are largely based on consideration of valence. Positive or negative affect drive the resulting triadic relations. The original consideration was for the cognitive balance of the individual, or the psychological congruity of affect by an individual for two objects that they see as causally connected. As such, the valenced relationship of any three actors are brought into combined consideration because any two are considered affectively combined through their joint combination as "the parts of a causal unit (Heider 1946:107)."

@@

In addition to these explanations, the distinction measure also reveals the utility of a distinction based theory of triadic closure. Figure 3 \textit{Model (a)} and Table 4 present a graph and corresponding measures of near-neighbor triadic closure. The distinction of nodes 1 and 6 increases to .106, compared to 0 in the circle graph, while all other nodes fall below 0 due to the reverberating effects of the new connection. Notably, node 7 increases in eigenvector status from the combination, but overall decrease in scaled distinction due to the tie. This example demonstrates the fragility of the circle graph as an equality condition in a distinction-based theoretical framework, as there are opportunities for mutual increases in distinction through triadic closure. While both eigenvector centrality and scaled distinction would predict tie formation in this simplified social structure, the scaled distinction measure would find that Node 7 would actually lose out in some ways from this arrangement. This matches with findings that often initial connections do weaken when triadic closure occurs \citep{huang2018will}, consistent with a reaction to the increased constraint placed on the closure broker. However, the actual incentives and consequences for each of the nodes in a triadic closure will be dependent on the overall existing network structure. Node 7 is worse off due to the structure of the circle graph, rather than because it will always be worse off under all conditions. If Node 7 is actually highly central to the graph, near neighbor triadic closure can be distinction increasing. Beyond questions of balancing valence as is the current framework for understanding triadic closure, a distinction-based theoretical framework can explain the propensity towards triadic closure, but also its failure to form depending on the initial patterning of relationships.

@@

What the difference between these two examples suggests is that, based on considerations for distinction, the initial spread of social organization occurs locally through the triadic balance of near neighbors. This would suggest that the natural form of organizational formation is the pyramid formed from combinations of local triads that are further connected by brokers or welds creating structural folds. However, the transitivity created by the initial triad formation would spread quickly throughout the network, making welding structural folds a highly advantageous strategy for members of triads who could facilitate that connection, which is further explored below. Additionally, the example of triadic closure shows that only mild additions to network ties can very quickly create stratification. The addition of a single connection between node 1 and 6 in Figure 3 \textit{Model (a)} stratified a network from equality into four levels of hierarchy with a single additional connection.

@@

%Circles as their own, pizza pie. at least get to 0 status. Wheel can emerge
%folds vs circles
\textit{Model (b)} presents an idealized depiction of a structural fold \citep{vedres2010structural}. The resulting measurement scores are in Table 2. While node 1 maintains high eigenvector centrality, the eigenvector centrality of the other nodes increases, while their constraint decreases. It is actually the constraint of Node 1 that increases, reaching 1.00 because all its neighbors are maximally interconnected in their respective subgroups. This position of brokerage is more constrained than the star graph because the other actors are connected in subgroups. The actors in the subgroup can combine their interests to counterbalance the independence of Node 1 in the center. In an unadjusted measure of distinction, the different nodes become remarkably close in their distinction measure. With a scalar adjustment, the stratification increases, giving Node 1 a slight advantage, but also highlighting the extent to which Node 1's distinction is eroded. In a maximally connected graph, the distinction of each node would also go to 0.00, as there is no network differentiation among nodes.

%brokerage and structural folds Figure 4
\begin{figure}[H]
    \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
    \centering
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/sf4.png}
            \caption{Structural Fold (4-Clique)}
            \label{fig:sf}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/sf5.png}
            \caption{Structural Fold (5-Clique)}
            \label{fig:sf}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/two-connected.png}
            \caption{Two-Connected}
            \label{fig:twoconnected}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/circle-weld.png}
            \caption{Circle Weld}
            \label{fig:circleweld}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/broker3.png}
            \caption{Brokerage Mediated (3-Cliques)}
            \label{fig:classicbrokerage3}
    \end{subfigure} 
    \label{fig:struct-fold}
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/broker4.png}
            \caption{Brokerage Mediated (4-Cliques)}
            \label{fig:classicbrokerage3}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/tree-mediator2.png}
            \caption{Tree with 2 Branches, 2 Leaves}
            \label{fig:treemediator2}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/tree-mediator3.png}
            \caption{Tree with 2 Branches, 3 Leaves}
            \label{fig:treemediator3}
    \end{subfigure}
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/tree-mediator2-long.png}
            \caption{Tree with 2 Branches, 2 Leaves, 1 Long Tie}
            \label{fig:treemediatorlong2}
    \end{subfigure}
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/tree-mediator3-long.png}
            \caption{Tree with 2 Branches, 3 Leaves, 1 Long Tie}
            \label{fig:treemediatorlong3}
    \end{subfigure}
    \caption{Comparison of betweenness brokerage and a structural fold goes here}
\end{figure}

>>>>

\citet[1263-4]{padgett1993robust} were the first to bring attention to how the cohesion of those around a broker can structure their own decisions. Cosimo de Medici, acting as a broker between two antagonistic groups, was constrained in the actions he could take because strong actions favoring one side would upset the other side. However, such a constraint was not an issue because Medici’s interests were in perpetuating his central role as broker, and, as the only source of interest, he was able to maintain the position. With regard to conceptualizing distinction, the theory of robust action puts forward that those at the center of political networks, as long as their interests derive solely from maintaining that position, engage in moderate strategies that do not overly antagonize either side. The decisions are based on ensuring their continued position at the center of the network.

@@

The second term $\kappa_i$ in equation~\ref{eq:distinction} is a \textit{localized} induced centrality score in the sense of \citet{borgatti2020oxford-fa6}. According to \citet{everett2010induced-817}, an induced centrality is obtained whenever we calculate the difference that removing a node in the graph makes to some global property of the graph (a graph invariant). \citet{everett2010induced-817} suggests taking the sum of the centrality scores of the nodes in the node-deleted subgraph produced by removing a given node as that node's contribution to the centrality of other nodes in the graph, as a measure of that node's centrality. They propose examples extending this idea to the classical centrality measures like degree, closeness, and betweenness \citep{freeman79}.

@@

Triadic closure is an observed regularity in which a connection is likely to form between two actors who, although not originally connected, are first connected to a common alter (Simmel 1908). In essence, friends of friends have an increased likelihood of also becoming friends. The causes of this regularity, and the reasons for its occasional failure, are largely based on consideration of valence. Positive or negative affect drives the resulting triadic relations. The original consideration was for the cognitive balance of the individual, or the psychological congruity of affect by an individual for two objects that they see as causally connected. As such, the valenced relationship of any three actors is brought into combined consideration because any two are considered affectively combined through their joint combination as ``the parts of a causal unit'' \citep[107]{heider1946attitudes-ff3}.

@@

In addition to these explanations, the distinction measure also reveals the utility of a distinction-based theory of triadic closure. Figure 3 \textit{Model (a)} and Table 4 present a graph and corresponding measures of near-neighbor triadic closure. The distinction of nodes 1 and 6 increases to .106, compared to 0 in the circle graph, while all other nodes fall below 0 due to the reverberating effects of the new connection. Notably, node 7 increases in eigenvector status from the combination, but overall decreases in scaled distinction due to the tie. This example demonstrates the fragility of the circle graph as an equality condition in a distinction-based theoretical framework, as there are opportunities for mutual increases in distinction through triadic closure. While both eigenvector centrality and scaled distinction would predict tie formation in this simplified social structure, the scaled distinction measure would find that Node 7 would actually lose out in some ways from this arrangement. This matches with findings that often initial connections do weaken when triadic closure occurs \citep{huang2018will}, consistent with a reaction to the increased constraint placed on the closure broker. However, the actual incentives and consequences for each of the nodes in a triadic closure will be dependent on the overall existing network structure. Node 7 is worse off due to the structure of the circle graph, rather than because it will always be worse off under all conditions. If Node 7 is actually highly central to the graph, near-neighbor triadic closure can be distinguished beyond questions of balancing valence, as is the current framework for understanding triadic closure. A distinction-based theoretical framework can explain the propensity towards triadic closure, but also its failure to form, depending on the initial patterning of relationships.

@@

What the difference between these two examples suggests is that, based on considerations for distinction, the initial spread of social organization occurs locally through the triadic balance of near neighbors. This suggests that the natural form of organizational formation is a pyramid formed from combinations of local triads, further connected by brokers or welds, which create structural folds. However, the transitivity created by the initial triad formation would spread quickly throughout the network, making welding structural folds a highly advantageous strategy for members of triads who could facilitate that connection, which is further explored below. Additionally, the example of triadic closure shows that only mild additions to network ties can very quickly create stratification. The addition of a single connection between node 1 and 6 in Figure 3 \textit{Model (a)} stratified a network from equality into four levels of hierarchy with a single additional connection.

@@

%Circles as their own, pizza pie. at least get to 0 status. Wheel can emerge
%folds vs circles
\textit{Model (b)} presents an idealized depiction of a structural fold \citep{vedres2010structural}. The resulting measurement scores are in Table 2. While node 1 maintains a high status score, the status of the other nodes increases, while their constraint decreases. It is actually the constraint of Node 1 that increases, reaching 1.00 because all its neighbors are maximally interconnected in their respective subgroups. This position of brokerage is more constrained than the star graph because the other actors are connected in subgroups. The actors in the subgroup can combine their interests to counterbalance the independence of Node 1 in the center. In an unadjusted measure of distinction, the different nodes become remarkably close in their distinction measure. With a scalar adjustment, the stratification increases, giving Node 1 a slight advantage, but also highlighting the extent to which Node 1's distinction is eroded. In a maximally connected graph, the distinction of each node would also go to 0.00, as there is no network differentiation among nodes.

%brokerage and structural folds Figure 4
\begin{figure}
    \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
    \centering
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/sf4.png}
            \caption{Structural Fold (4-Clique)}
            \label{fig:sf}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/sf5.png}
            \caption{Structural Fold (5-Clique)}
            \label{fig:sf}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/two-connected.png}
            \caption{Two-Connected}
            \label{fig:twoconnected}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/circle-weld.png}
            \caption{Circle Weld}
            \label{fig:circleweld}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/broker3.png}
            \caption{Brokerage Mediated (3-Cliques)}
            \label{fig:classicbrokerage3}
    \end{subfigure} 
    \label{fig:struct-fold}
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/broker4.png}
            \caption{Brokerage Mediated (4-Cliques)}
            \label{fig:classicbrokerage3}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/tree-mediator2.png}
            \caption{Tree with 2 Branches, 2 Leaves}
            \label{fig:treemediator2}
    \end{subfigure} 
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/tree-mediator3.png}
            \caption{Tree with 2 Branches, 3 Leaves}
            \label{fig:treemediator3}
    \end{subfigure}
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/tree-mediator2-long.png}
            \caption{Tree with 2 Branches, 2 Leaves, 1 Long Tie}
            \label{fig:treemediatorlong2}
    \end{subfigure}
    \begin{subfigure}[b]{0.23\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/tree-mediator3-long.png}
            \caption{Tree with 2 Branches, 3 Leaves, 1 Long Tie}
            \label{fig:treemediatorlong3}
    \end{subfigure}
    \caption{Comparison of betweenness brokerage and a structural fold goes here}
\end{figure}
